A Mechanised Library for Generic (Co)Programming

### 2024-01-31 paper/paper.tex
Recursion schemes from comonads

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
\newcommand{\hylo}[2]{\cata{#1 \to #2}}
\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\head}{\operator{head}}
\newcommand{\tail}{\operator{tail}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}
\newcommand{\fmap}[1]{\text{fmap}\;#1}

@@

\paragraph{Free Monads}
Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
solution to the following domain equation
\[
  D^{*}A = A + DD^{*}A
\]
where $\eta_{A} : A \to D^{*}A$ is the left injection and
$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
$a : DA \to A$ there is a unique solution to the following recursive equation
\[
  x \comp \call_{A} = a \comp D(x)
\]
that is $\catafree{a} = \cata{[id_{A}; a]}$.

\paragraph{Cofree Comonads}
Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
greatest solution to the following domain equation
\[
  D_{\infty}A = A + DD_{\infty}A
\]
where $\epsilon_{A} : D_{\infty}A \to A$ is the first projection and
$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.

Given an algebra $c : A \to DA$ there is a unique solution to the following
recursive equation
\[
  \tail_{A} \comp x = D(x) \comp c
\]
that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$.

@@

One option is to prove that the following function (given by a catamorphism) is
equivalent to the above
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} (0) = (\top, \bot)\\
  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
\end{align*}
This function is clearly a catarmorphism:
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} = \cata{a} \operator{ where }\\
  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
  & \qquad \qquad a (\operator{nil}) = (\top, \bot)\\
  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
\end{align*}
where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
the associated recursive function.

@@

\subsection{Course-of-Value Recursion}
First we have to assume a distributive law $\lambda : DG \to GD$.
For an algebra $a : D(D_{\infty} A) \to A$,
\begin{align*}
  & g : \mu F \to A\\
  & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}
where $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX  $ is a $D$-algebra on $GX$. The catamorphism of this algebra has the following type
\[
  \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
\]
which is the type of a $G$-algebra used to initialise the memory table.

\paragraph{Hylo Implementation}
Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
what is the unique solution of the following equation
\begin{align*}
  & f : \mu F \to G_{\infty} A\\
  & f = a \comp \fmap{f} \comp \InOp
\end{align*}
The solution to this equation can be found again using an hylo
$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.
The course-of-value recursion can be recovered from this by setting
$g = \head \comp f$

>>>>

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
\newcommand{\hylo}[2]{\cata{#1 \to #2}}
\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\head}{\operator{head}}
\newcommand{\inl}{\operator{inl}}
\newcommand{\inr}{\operator{inr}}
\newcommand{\tail}{\operator{tail}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}
\newcommand{\fmap}[1]{\text{fmap}\;#1}

\newcommand{\zero}{\operator{zero}}
\newcommand{\Nil}{\operator{nil}}
\newcommand{\Succ}{\operator{succ}}

@@

\paragraph{Free Monads}
Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
solution to the following domain equation
\[
  D^{*}A = A + DD^{*}A
\]
where $\eta_{A} : A \to D^{*}A$ is the left injection and
$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
$a : DA \to A$ there is a unique solution to the following recursive equation
\[
  x \comp \call_{A} = a \comp D(x)
\]
that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
defined as $\mu = \catafree{\call}$.

\paragraph{Cofree Comonads}
Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
greatest solution to the following domain equation
\[
  D_{\infty}A = A + DD_{\infty}A
\]
where $\head : D_{\infty}A \to A$ is the first projection and
$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
Given an algebra $c : A \to DA$ there is a unique solution to the following
recursive equation
\[
  \tail_{A} \comp x = D(x) \comp c
\]
that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
can extend $\tail$ to the comultiplication for the comonad
$\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
$\delta = \anacofree{\tail}$.

@@

One option is to prove that the following function (given by a catamorphism) is
equivalent to the above
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} (0) = (\top, \bot)\\
  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
\end{align*}
This function is clearly a catarmorphism:
\begin{align*}
  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
  & \operator{even-odd} = \cata{a} \operator{ where }\\
  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
  & \qquad \qquad a (\Nil) = (\top, \bot)\\
  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
\end{align*}
where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
the associated recursive function.

@@

\subsection{Course-of-Value Recursion}
The knapsack algorithm is defined as following in pseudo-Haskell code:
\begin{align*}
  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
  & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
    [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
\end{align*}
%
with dynamic programming knapsack can be rephrased like this
\begin{align*}
  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
  & \operator{knapsack}\;wvs\;c= table !! c \\
  & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
  & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
\end{align*}

The recursion scheme is justified by solving a recursive equation.  First we
define an operation to create a memory table representing the list of capacities
$[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
for constructing the naturals $\Nat = \mu D$ the we define the generating
function by induction on the naturals
\begin{align*}
  & \operator{fan} : \Nat \to D_{\infty}\Nat\\
  & \operator{fan} (0)    = \pair{0}{\inl}\\
  & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
\end{align*}
The recursion scheme from a comonad is then defined as follows
\begin{align*}
  & \operator{rsfc} : \Nat \to A\\
  & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
\end{align*}
At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
\begin{align*}
  & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
  & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
  & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
  & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
\end{align*}
and now we can define the algebra over the reals
\begin{align*}
  & \operator{alg} : DD_{\infty} \R \to \R\\
  & \operator{alg}\; (\inl)  = 0\\
  & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
\end{align*}

It remains to show how an $\operator{rsfc}$ is implemented in terms of hylomorhisms.

% A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$

% For example, for the functor $(1 + -)$ an algebra over this functor is a
% function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
% $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
% \[
%   D_{\infty} A \iso A \times (1 + D_{\infty})
% \]
% which is the the type of non-empty possibly infinite lists with elements in $A$

% First we have to assume a distributive law
% $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
% $a : D(D_{\infty} A) \to A$,
% \begin{align*}
%   & g : \mu D \to A\\
%   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
% \end{align*}
% where
% $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
% is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
% type
% \[
%   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
% \]
% which is the type of a $G$-algebra used to initialise the memory table.

% Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
% what is the unique solution of the following equation
% \begin{align*}
%   & f : \mu F \to G_{\infty} A\\
%   & f = a \comp \fmap{f} \comp \InOp
% \end{align*}
% The solution to this equation can be found again using an hylo
% $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
% can be recovered from this by setting $g = \head \comp f$

% \subsubsection{Histomorphism}

% The definition of $\lambda$ is:
% \begin{align*}
%   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
%   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% \end{align*}
% For example, for $GX = 1 + X$ we have to define
% \begin{align*}
%   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
%   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
%   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% \end{align*}

% % \begin{align*}
% %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % \end{align*}

% % Defining the algebra
% % \begin{align*}
% %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% %   & \operator{alg}\; Zero           = 0\\
% %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
% % \end{align*}
